### 2024-11-04 draft-2/draft-2.tex
added provisional description of notation

@@

  # Pseudocode
  amber_fg = lch(0.7, 0.125,  75)
  green_fg = lch(0.6, 0.15 , 150)
   blue_fg = lch(0.5, 0.2  , 270)
  lilac_fg = lch(0.6, 0.2  , 300)
\endruby
%
\newtoks\restorecolor \restorecolor={0 0 0}%
%
\long\def\color#1#2{%
  {\restorecolor={#1}\special{color rgb #1}#2}%
  \special{color rgb \the\restorecolor}%
}%
%
\long\def\colorbox#1#2#3{%
  \par
  % Set the contents of the box
  \advance\hsize by -2em
  \setbox0 = \vbox{#3}%
  \advance\hsize by 2em
  % Insert the usual interline glue
  \ifdim \prevdepth > -1000pt
    \vskip-\prevdepth
    \vskip\baselineskip
  \fi
  % Insert an extra half of a baselineskip above
  \vskip0.5\baselineskip
  % Calculate dimensions
  \dimen0 = \ht0
  \dimen1 = \dp0
  \advance\dimen0 by 0.25\baselineskip % midway along half-baselineskip above
  \advance\dimen0 by 0.25\baselineskip % adjust the upper edge and
  \advance\dimen1 by 0.20\baselineskip %   lower edge to be perceptually correct
  \advance\dimen1 by 0.25\baselineskip % midway along half-baselineskip below
  % Insert the background
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#1}{\hrule width \hsize height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the ribbon
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#2}{\hrule width 0.5em height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the box
  \hbox{\hskip1em\box0}
  % Insert an extra half of a baselineskip below
  \vskip0.5\baselineskip
}%
\long\def  \greybox#1{\colorbox{\r{  grey_bg}}{\r{  grey_rbn}}{#1}}%
\long\def\purplebox#1{\colorbox{\r{purple_bg}}{\r{purple_rbn}}{#1}}%
\long\def  \bluebox#1{\colorbox{\r{  blue_bg}}{\r{  blue_rbn}}{#1}}%
\long\def \greenbox#1{\colorbox{\r{ green_bg}}{\r{ green_rbn}}{#1}}%
\long\def\yellowbox#1{\colorbox{\r{yellow_bg}}{\r{yellow_rbn}}{#1}}%
\long\def   \redbox#1{\colorbox{\r{   red_bg}}{\r{   red_rbn}}{#1}}%
%
\long\def\label  #1{\color{\r{ label_fg}}{#1}}%
\long\def\term   #1{\color{\r{purple_fg}}{#1}}%
\long\def\comment#1{\color{\r{  grey_fg}}{{\noto ※ }#1}}%
\long\def\s      #1{\color{\r{  teal_fg}}{\fw #1}}%
%
\def\hex#1{{\fw\color{\r{ltgrey_fg}}{x}#1}}%
\def\uni#1{{\caps u+#1}}%
%
\def\ellipsis{.\ot.\ot.}
%
\newcount\majorcount
\newcount\minorcount
\def\major{%
  \advance\majorcount by 1
  \minorcount=0
  \leavevmode
  \llap{\label{\the\majorcount}\hem}%
}%
\def\minor{%
  \advance\minorcount by 1
  \leavevmode
  \llap{\label{\the\majorcount·\the\minorcount}\hem}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gets{<-}%
\def\to  {->}%
\def\then{=>}%
%
\setbox0=\hbox{\fw ()[]\char"7B\char"7D\it ()[]\char"7B\char"7D}%
\newdimen\strutht
\newdimen\strutdp
\strutht=\ht0
\strutdp=\dp0
\dimen0=\baselineskip
\advance\dimen0 by -\strutht
\advance\dimen0 by -\strutdp
\divide\dimen0 by 2
\advance\strutht by \dimen0
\advance\strutdp by \dimen0
\def\strut{\vrule width 0pt height \strutht depth \strutdp}%
%
\def\tab{%
  \kern 0.75ex
  \color{\r{ltgrey_fg}}{\vrule width 0.5pt}%
  \kern -0.5pt
  \kern -0.75ex
  \kern 3ex
}%
%
\def\kwd  #1{\color{\r{  blue_fg}}{#1}}% def, return, break, error
\def\flow #1{\color{\r{ amber_fg}}{#1}}% repeat, if, then, match
\def\liter#1{\color{\r{   red_fg}}{#1}}% numeric
\def\const#1{\color{\r{ lilac_fg}}{#1}}% empty, eof
\def\func #1{\color{\r{ green_fg}}{#1}}%
\def\var  #1{{\ifw #1}}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\vglue1\baselineskip
\centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
\centerline{\lcap4 october 2024 • draft}

\vskip4\baselineskip

@@

This specification governs the interaction between two processes named the
{\it client} and the {\it engine}. \comment{Graphical interfaces, terminal
emulators, and scripts and utilities are examples of clients.} The text of
sections 1–4 is normative (except as described in {\it Conventions} below).

@@

{\bf Conventions}

Sequences are written with angle brackets; for example, ⟨\ot0, 1, 2,
\ellipsis\pt⟩. An~{\fw\color{\r{ltgrey_fg}}{x}} preceding a number
indicates that it is written in hexadecimal; for example, \hex{10} = 16.
Sequences of bytes encoded in {\caps ascii} are set in teal; for example,
\s{uci} = ⟨\ot\hex{75}, \hex{63}, \hex{69}⟩\pt. Sequences of tokens may be
written without commas when they contain token literals; for example,
⟨\ot\s{ab}~\s{c}~{\it d}~\s{e}\ot⟩ rather than
⟨\ot\s{ab},~\s{c},~{\it d},~\s{e}\ot⟩.

Special terms defined by the specification are set in \term{purple} when they
are first introduced and inline comments are set \comment{after a reference
mark}.
\vskip-\parskip
\bluebox{A blue box is used to describe a convention that clients and engines
are encouraged to follow, usually to do with the interpretation or meaning of
the messages that clients and engines send. A blue box is also used to provide
a recommendation, usually to do with implementation-defined behavior.}
\vskip-\parskip
\greybox{A grey box is used to provide an explanative note or comment.}
\vskip-\parskip
\greenbox{A green box is used to provide an example of conforming behavior.}
\vskip-\parskip
Text within colored boxes is nonnormative.

@@

\minor A \term{Forsyth–Edwards Notation ({\caps fen}) record} is a sequence of
six tokens ⟨{\it board, side to move, rights, {\icap ep} target, {\icap dfz},
move number}⟩ of the form\par
%
\dimen0=5em
\vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
  \dt=\dt\ \r{(['{\it row}']*8).join('\pt\s/\pt')}\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
  \dt=\dt\ \s8 \alt\ [\s1–\s7\s{KQRBNPkqrbnp}]+\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it side to move}
  \dt=\dt\ \s w \alt\ \s b\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it rights}
  \dt=\dt\ \s- \alt\ \s K?\pt\s Q?\pt\s k?\pt\s q?\par
  % \lower0.5pt\hbox{\notomono ∩} [\s{KQkq}]+)\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it {\icap ep} target}
  \dt=\dt\ \s- \alt\ [\s a–\s h][\s{36}]\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]?\ \alt\ \s{100}\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
  \dt=\dt\ [\s1–\s9][\s0–\s9]\char"7B\ot0,\pt3\ot\char"7D\par
\vskip-\parskip\hskip3ex
  \comment{The dash “\s-” is \hex{2d}.}

@@

\minor The client must enqueue a \s{ucinewgame} message at least once before
before it enqueues a \s{go} message for the first time after the initial state.
\vskip-\parskip
\bluebox{Engines are strongly recommended to behave identically whether or not
a \s{ucinewgame} message is ever sent.}

\minor The client must enqueue a \s{position} message at least once before it
enqueues a \s{go} message for the first time after the initial state.
\vskip-\parskip
\bluebox{Engines are strongly recommended, if a \s{position} message is not
sent before the first \s{go} message, to behave as if \s{position startpos}
were sent.}
\vskip-\parskip
\greybox{Note that the client is not required to send \s{isready} before
\s{go}.}
\vskip-\parskip
\greybox{Note that the client may send \s{ucinewgame} after \s{position}
before \s{go} without sending any additional intervening \s{position} message.}
\vskip-\parskip
\greybox{Note that \s{ucinewgame} does not reset the current board state nor
does it reset the move history. (Only \s{position} changes the current board
state and move history.)}

\minor The specification ceases to govern the interaction between the client
and the engine when the client enqueues a well-formed \s{quit} message.
\vskip-\parskip
\bluebox{Engines are recommended to stop searching immediately and exit.
Clients are recommended, if engines are their subprocesses, to provide a
grace period before terminating engines.}

% \major{\bf Notation}

>>>>

\font\cas=\ldfont{Cascadia Mono Italic SemiLight Italic}{9.5}{\codefeat}%

@@

  # Pseudocode
  amber_fg = lch(0.7, 0.125,  75)
  green_fg = lch(0.6, 0.15 , 150)
   blue_fg = lch(0.5, 0.2  , 270)
  lilac_fg = lch(0.6, 0.2  , 300)
\endruby
%
\newtoks\restorecolor \restorecolor={0 0 0}%
%
\long\def\color#1#2{%
  {\restorecolor={#1}\special{color rgb #1}#2}%
  \special{color rgb \the\restorecolor}%
}%
%
\long\def\colorbox#1#2#3{%
  \par
  % Set the contents of the box
  \advance\hsize by -2em
  \setbox0 = \vbox{#3}%
  \advance\hsize by 2em
  % Insert the usual interline glue
  \ifdim \prevdepth > -1000pt
    \vskip-\prevdepth
    \vskip\baselineskip
  \fi
  % Insert an extra half of a baselineskip above
  \vskip0.5\baselineskip
  % Calculate dimensions
  \dimen0 = \ht0
  \dimen1 = \dp0
  \advance\dimen0 by 0.25\baselineskip % midway along half-baselineskip above
  \advance\dimen0 by 0.25\baselineskip % adjust the upper edge and
  \advance\dimen1 by 0.20\baselineskip %   lower edge to be perceptually correct
  \advance\dimen1 by 0.25\baselineskip % midway along half-baselineskip below
  % Insert the background
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#1}{\hrule width \hsize height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the ribbon
  \kern -0.25\baselineskip % midway along extra
  \kern -0.25\baselineskip % adjustment
  \color{#2}{\hrule width 0.5em height \dimen0 depth \dimen1}%
  \kern -\ht0
  \kern -\dp0
  \kern -0.20\baselineskip % adjustment
  \kern -0.25\baselineskip % midway along extra
  % Insert the box
  \hbox{\hskip1em\box0}
  % Insert an extra half of a baselineskip below
  \vskip0.5\baselineskip
}%
\long\def  \greybox#1{\colorbox{\r{  grey_bg}}{\r{  grey_rbn}}{#1}}%
\long\def\purplebox#1{\colorbox{\r{purple_bg}}{\r{purple_rbn}}{#1}}%
\long\def  \bluebox#1{\colorbox{\r{  blue_bg}}{\r{  blue_rbn}}{#1}}%
\long\def \greenbox#1{\colorbox{\r{ green_bg}}{\r{ green_rbn}}{#1}}%
\long\def\yellowbox#1{\colorbox{\r{yellow_bg}}{\r{yellow_rbn}}{#1}}%
\long\def   \redbox#1{\colorbox{\r{   red_bg}}{\r{   red_rbn}}{#1}}%
%
\long\def\label  #1{\color{\r{ label_fg}}{#1}}%
\long\def\term   #1{\color{\r{purple_fg}}{#1}}%
\long\def\comment#1{\color{\r{  grey_fg}}{{\noto ※ }#1}}%
\long\def\s      #1{\color{\r{  teal_fg}}{\fw #1}}%
%
\def\hex#1{{\fw\color{\r{ltgrey_fg}}{\cas x}#1}}%
\def\uni#1{{\caps u+#1}}%
%
\def\ellipsis{.\ot.\ot.}
%
\newcount\majorcount
\newcount\minorcount
\def\major{%
  \advance\majorcount by 1
  \minorcount=0
  \leavevmode
  \llap{\label{\the\majorcount}\hem}%
}%
\def\minor{%
  \advance\minorcount by 1
  \leavevmode
  \llap{\label{\the\majorcount·\the\minorcount}\hem}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gets{<-}%
\def\to  {->}%
\def\then{=>}%
%
\setbox0=\hbox{\fw ()[]\char"7B\char"7D\it ()[]\char"7B\char"7D}%
\newdimen\strutht
\newdimen\strutdp
\strutht=\ht0
\strutdp=\dp0
\dimen0=\baselineskip
\advance\dimen0 by -\strutht
\advance\dimen0 by -\strutdp
\divide\dimen0 by 2
\advance\strutht by \dimen0
\advance\strutdp by \dimen0
\def\strut{\vrule width 0pt height \strutht depth \strutdp}%
%
\def\tab{%
  \kern 0.75ex
  \color{\r{ltgrey_fg}}{\vrule width 0.5pt}%
  \kern -0.5pt
  \kern -0.75ex
  \kern 3ex
}%
%
\def\kwd  #1{\color{\r{  blue_fg}}{#1}}% def, return, break, error
\def\flow #1{\color{\r{ amber_fg}}{#1}}% repeat, if, then, match
\def\liter#1{\color{\r{   red_fg}}{#1}}% numeric
\def\const#1{\color{\r{ lilac_fg}}{#1}}% empty, eof
\def\func #1{\color{\r{ green_fg}}{#1}}%
\def\var  #1{{\ifw #1}}%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\centerline{\lcap8 UNIVERSAL CHESS INTERFACE}
\centerline{\lcap4 october 2024 • draft}

\vskip3\baselineskip

@@

This specification governs the interaction between two processes named the
{\it client} and the {\it engine}. \comment{Graphical interfaces, terminal
emulators, and scripts and utilities are examples of clients.} The text of
sections 1–4 is normative (except within comments and colored boxes, described
in the section {\it Notation} below).

@@

{\bf Notation}

An~\hex{} preceding a number indicates that it is written in hexadecimal; for
example, \hex{10} = 16. A letter or word set in teal is an abbreviation for a
byte or sequence of bytes encoded in {\caps ascii}; for example,
\s{k} = \hex{6b} and \s{uci} = ⟨\ot\hex{75}, \hex{63}, \hex{69}⟩\pt. Special
terms defined by the specification are set in \term{purple} when they are
first introduced and inline comments are set \comment{after a reference
mark}. Comments are nonnormative.
\vskip-\parskip
\bluebox{A blue box is used to describe a convention that clients and engines
are encouraged to follow, usually to do with the interpretation or meaning of
the different messages that clients and engines can send. A blue box is also
used to provide a recommendation, usually to do with implementation-defined
behavior.}
\vskip-\parskip
\greybox{A grey box is used to provide an explanative note or comment.}
\vskip-\parskip
\greenbox{A green box is used to provide an example of conforming behavior.}
\vskip-\parskip
Text within colored boxes is nonnormative.

Angle brackets are sometimes used to delimit sequences and commas sometimes used
to separate items in a sequence. For example, ⟨\ot\s{go}~\s{depth} {\it d}
\s{movetime}~\s{1000}\ot⟩ and
⟨\ot\s{go},~\s{depth},~{\it d},~\s{movetime}~\s{1000}\ot⟩ are equivalent, as
are \s{go depth 8 movetime 1000} and ⟨\ot\s{go}, \s{depth}, \s{8}, \s{movetime},
\s{1000}\ot⟩.

\def\synitem#1#2{\line{\hskip3ex\hbox to 6em{#1\hfil}\em #2\hfil}}%

Patterns for sequences of bytes are written using traditional syntax for regular
expressions:\par
\synitem{parentheses}{group}
\synitem{brackets}{byte class}
\synitem{en dash}{byte range}
\synitem{vertical bar}{alternation}
\synitem{plus sign}{repetition (once or more)}
%\synitem{asterisk}{repetition (zero times or more)}
\synitem{question mark}{repetition (zero times or once)}
\synitem{braces}{repetition ({\it n} times, or at least {\it n₁} and at most {\it n₂})}

Patterns for sequences of tokens are written using analogous syntax.

@@

\minor A \term{Forsyth–Edwards Notation ({\caps fen}) record} is a sequence of
six tokens ⟨{\it board, side to move, rights, {\icap ep} target, {\icap dfz},
move number}⟩ of the form\par
%
\dimen0=5em
\vskip-0.5\parskip\hskip3ex\rjust{\dimen0}{\it board}
  \dt=\dt\ {\it row}\pt(\s/\pt{\it row}\pt)\char"7B\ot7\ot\char"7D\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it row}
  \dt=\dt\ \s8 \alt\ [\s1–\s7\s{KQRBNPkqrbnp}]+\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it side to move}
  \dt=\dt\ \s w \alt\ \s b\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it rights}
  \dt=\dt\ \s- \alt\ \s K?\pt\s Q?\pt\s k?\pt\s q?\par
  % \lower0.5pt\hbox{\notomono ∩} [\s{KQkq}]+)\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it {\icap ep} target}
  \dt=\dt\ \s- \alt\ [\s a–\s h][\s{36}]\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\icap dfz}
  \dt=\dt\ \s0 \alt\ [\s1–\s9][\s0–\s9]?\ \alt\ \s{100}\par
\vskip-\parskip\hskip3ex\rjust{\dimen0}{\it move number}
  \dt=\dt\ [\s1–\s9][\s0–\s9]\char"7B\ot0,\pt3\ot\char"7D\par
\vskip-\parskip\hskip3ex
  \comment{The dash “\s-” is \hex{2d}.}

@@

\minor The client must enqueue a \s{ucinewgame} message at least once before
before it enqueues a \s{go} message for the first time after enqueuing \s{uci}.
\vskip-\parskip
\bluebox{Engines are strongly recommended to behave identically whether or not
a \s{ucinewgame} message is ever sent.}

\minor The client must enqueue a \s{position} message at least once before it
enqueues a \s{go} message for the first time after enqueuing \s{uci}.
\vskip-\parskip
\bluebox{Engines are strongly recommended, if a \s{position} message is not
sent before the first \s{go} message, to behave as if \s{position startpos}
were sent.}
\vskip-\parskip
\greybox{Note that the client is not required to send \s{isready} before
\s{go}.}
\vskip-\parskip
\greybox{Note that the client may send \s{ucinewgame} after \s{position}
before \s{go} without sending any additional intervening \s{position} message.}
\vskip-\parskip
\greybox{Note that \s{ucinewgame} does not reset the current board state nor
does it reset the move history. (Only \s{position} changes the current board
state and move history.)}

\minor The specification ceases to govern the interaction between the client
and the engine when the client enqueues a well-formed \s{quit} message.
\vskip-\parskip
\bluebox{Engines are recommended to stop searching immediately and exit.
Clients are recommended, if the engine is their subprocess, to provide a
grace period before terminating the engine.}
